---
title: "Has the Composition of Saudi Exports Shifted?"
author:  "Abdullah Adel Al-Madhi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    highlight: pygments
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    code_folding: show
    df_print: kable
    fig_caption: true
    self_contained: true
---

# 1. Introduction
 
Saudi Arabia has been working to reduce its dependence on raw material
exports as part of **Vision 2030**. The usual way to measure this is to
look at total export values, but this approach has a problem: export
values move with **oil prices** and **inflation**, so a rise in value
does not necessarily mean a change in structure.
 
This report asks a different question:
 
> **Has the composition of Saudi exports shifted from raw materials
> toward manufactured goods between 2015 and 2026?**
 
To answer it we look at four things:
 
1.  **What the composition looks like** and how it changed over the
    period.
 
2.  **Whether the change is real growth** or only a side effect of
    falling oil exports.
 
3.  **Whether the observed trend is statistically significant** rather
    than random fluctuation.
 
4.  **Where the trend is heading** by 2030, and how much confidence we
    can place in that forecast.
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4, fig.align = "center")
```
 
# 2. The Data
 
The data comes from the **Statistical Database** of the General Authority
for Statistics. It covers monthly export values from **January 2015 to
May 2026**, which gives us **137 months**. Exports are classified by
nature of materials into three categories:
 
1. **Manufactured** — finished industrial products, the highest
   value-added category.
 
2. **Semi-manufactured** — partially processed goods such as
   petrochemicals and refined products.
 
3. **Raw materials** — unprocessed exports, mainly crude oil.
 
These three categories are not arbitrary. They form a **value-added
ladder**, and moving up this ladder is exactly what economic
diversification means in practice.
 
## 2.1 Preparing the data
 
The original file was not ready for analysis and needed three steps:
 
1. The data came at the **transaction level**, meaning every single
   export operation had its own row. This gave **1,611,855 rows** in
   total, so we aggregated them by category and month. The result is
   **411 observations**, which is 137 months for each of the three
   categories.
 
2. The **date format** was not consistent. January was written as
   `YYYY / MM` while every other month used `M/D/YY`, so both formats
   had to be converted before the series could be ordered in time.
 
3. The **export values** were stored as text with thousand separators,
   so they had to be converted to numeric.
 
```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)     # Graphs
library(forecast)    # Arima & ggAcf & AutoPlot
library(tseries)     # ADF Test
library(patchwork)   # Merge two Graphs
library(sandwich)    # standard errors Correction
library(lmtest)      # coeftest
library(kableExtra)  # Summary tables
 
theme_set(theme_minimal(base_size = 11) +
          theme(plot.background  = element_rect(fill = "#fdf6e3", color = NA),
                panel.background = element_rect(fill = "#fdf6e3", color = NA),
                panel.grid.major = element_line(color = "#eadfc0", linewidth = 0.4),
                panel.grid.minor = element_blank()))
```
 
```{r}
# Clean file
monthly = read.csv("exports_clean.csv", fileEncoding = "UTF-8")
 
monthly$date     = as.Date(monthly$date)
monthly$category = trimws(monthly$category)
 
# Category labels: Arabic -> English
lab = c("مواد خام"  = "Raw",
        "نصف مصنعة" = "Semi-manufactured",
        "مصنعة"     = "Manufactured")
 
monthly$cat = unname(lab[monthly$category])
 
monthly$cat = factor(monthly$cat,
  levels = c("Raw", "Semi-manufactured", "Manufactured"))
 
stopifnot(!any(is.na(monthly$cat)))
 
monthly = monthly[order(monthly$date, monthly$cat), ]
rownames(monthly) = NULL
 
# Year, for the yearly comparisons
monthly$year = as.numeric(format(monthly$date, "%Y"))
 
# Share = category value / total for the same month
monthly$share = 100 * monthly$value / ave(monthly$value, monthly$date, FUN = sum)
 
# Monthly total, in SAR million
tot = tapply(monthly$value, monthly$date, sum)
 
man   = monthly[monthly$cat == "Manufactured", ]
man   = man[order(man$date), ]
man$t = 1:nrow(man)   # 1 = January 2015 ... 137 = May 2026
 
stopifnot(all(c("date", "cat", "year", "share", "value") %in% names(monthly)))
 
cat("rows:", nrow(monthly), "| months:", length(unique(monthly$date)), "\n")
cat("range:", format(range(monthly$date)), "\n")
```
 
## 2.2 Working with shares
 
Since the question is about **composition** and not about volume, we do
not use the export values directly. For each month we compute the
**share** of each category:
 
$$s_{i,t} = \frac{V_{i,t}}{\sum_{j} V_{j,t}} \times 100$$
 
where $V_{i,t}$ is the export value of category $i$ in month $t$, and
$i \in \{\text{Manufactured}, \text{Semi-manufactured}, \text{Raw}\}$.
 
By construction the three shares satisfy:
 
$$\sum_{i} s_{i,t} = 100 \qquad \text{for every month } t$$
 
This matters because absolute export values move with **oil prices** and
**inflation**. A year where oil prices double would show a large rise in
total exports without any real change in what the country actually
produces. To see why the share removes that effect, suppose every
category's value is multiplied by a common factor $\lambda_t > 0$:
 
$$
\tilde{s}_{i,t}
= \frac{\lambda_t V_{i,t}}{\sum_j \lambda_t V_{j,t}} \times 100
= s_{i,t}
$$
 
The share is invariant to any common scaling of the three series. What
survives is the structure itself.
 
We return to this choice in Section 10, because working with shares also
introduces a constraint that affects how the results should be read.
 
# 3. Descriptive Statistics
 
Before examining how the composition changed over time, it is worth
seeing what each category looks like across the whole period. The tables
below report the five-number summary, the mean and the standard
deviation — first for each category's share of total exports, then for
its value in absolute terms. All values are in millions of Saudi riyals
(SAR million); total monthly exports averaged
**`r format(round(mean(tot)), big.mark = ",")`** SAR million over the
period, or roughly
**`r sprintf("%.2f", 12 * mean(tot) / 1e6)`** trillion riyals a year.
 
```{r}
stats = function(x) c(summary(x), "Standard Deviation" = sd(x))
 
round(sapply(split(monthly$share, monthly$cat), stats), 2)
round(sapply(split(monthly$value, monthly$cat), stats), 1)
```
 
Four points stand out.
 
1. **Raw materials dominate the composition.** Their share averages
   **62.1%**, roughly three times the manufactured share (**20.5%**) and
   three and a half times the semi-manufactured share (**17.4%**). Even
   in its weakest month the raw share held **50.5%**, so raw materials
   never fell below half of total exports.
 
2. **The share series are skewed in opposite directions.** For raw
   materials the mean (**62.09**) falls below the median (**62.90**), and
   the distance from the minimum to the first quartile (**8.15** points)
   exceeds the distance from the third quartile to the maximum (**5.58**
   points): the long tail lies on the low side. The other two categories
   show the mirror image — manufactured has a mean (**20.51**) above its
   median (**19.37**), with an upper tail of **11.82** points against a
   lower tail of only **4.36**.
 
3. **Dispersion is not the same as instability.** Measured on shares, raw
   materials have the largest standard deviation (**4.48**), slightly
   above manufactured (**4.22**), with semi-manufactured the most stable
   (**3.02**). Measured on values, the ranking reverses: relative to its
   own mean, manufactured exports vary the most (6900.8 / 17737.8 ≈
   **39%**), raw materials less (**28%**), and semi-manufactured least
   (**20%**).
 
4. **The manufactured series covers a wide range.** Its share runs from
   **13.5%** to **33.7%**, a spread of 20 points, and its value from
   **6,787** to **34,256** SAR million — a factor of five between the
   weakest and the strongest month.
 
Taken together, the skew running in one direction for raw materials and
in the opposite direction for the other two categories suggests the
extreme months are not scattered at random, but concentrated at one end
of the period. The next section examines that directly.
 
# 4. How the Composition Changed
 
```{r, fig.width=8, fig.height=4.5}
pal = c("Raw"               = "#1baf7a",
        "Semi-manufactured" = "#eb6834",
        "Manufactured"      = "#2a78d6")
ends = monthly[monthly$date == max(monthly$date), ]
 
ggplot(monthly, aes(date, share, color = cat)) +
  geom_line(linewidth = 0.9) +
  geom_text(data = ends, aes(label = sprintf("%s  %.1f%%", cat, share)),
            hjust = 0, nudge_x = 60, size = 3.2, fontface = "bold",
            show.legend = FALSE) +
  scale_color_manual(values = pal) +
  scale_x_date(expand = expansion(mult = c(0.02, 0.28))) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Composition of Saudi Exports by Nature of Materials",
       subtitle = "Monthly share of total exports, January 2015 - May 2026",
       x = NULL, y = NULL) +
  theme(legend.position = "none",
        plot.title      = element_text(face = "bold", size = 13),
        plot.subtitle   = element_text(color = "grey40", size = 9.5),
        axis.text       = element_text(color = "grey35"))
```
 
The chart shows the answer to our first question. The **manufactured**
share starts at **`r sprintf("%.1f%%", man$share[1])`** in January 2015
and ends at **`r sprintf("%.1f%%", man$share[nrow(man)])`** in May 2026,
while the **raw materials** share moves in the opposite direction. The
**semi-manufactured** share changes very little.
 
Two features of the chart deserve attention:
 
1. **The movement is not smooth.** All three series fluctuate month to
   month, and there are sharp reversals in 2016, 2020 and 2023. We come
   back to these in Section 5.4.
 
2. **The rise is not constant.** The manufactured share climbs quickly
   until 2016, stays roughly flat between 2017 and 2020, then rises again
   and accelerates toward the end of the period.
 
To put numbers on what the chart shows, we compare the average share in
the first full year with the last full year.
 
## 4.1 Comparing shares: 2015 against 2025
 
```{r}
aggregate(share ~ cat + year,
          data = monthly[monthly$year %in% c(2015, 2025), ],
          FUN = mean)
```
 
 The two years are ten apart and the contrast is sharp. The manufactured
share nearly doubled, from **16.5%** to **29.1%**, a gain of **12.6**
percentage points. Raw materials gave up almost exactly that amount,
falling **10.8** points from **65.4%** to **54.6%**, while the
semi-manufactured share edged down **1.7** points. The adding-up
constraint guarantees that these three changes sum to zero, so the
question is not whether one category rose at another's expense — it
must have — but whether the rise reflects real expansion. The next
section answers that.


## 4.2 Checking the absolute values
 
A share can rise for two very different reasons: the numerator grew, or
the denominator shrank. The question matters, because only the first is
diversification. Writing the manufactured share in full makes the
ambiguity explicit:
 
$$
s_{\text{Man},t} = \frac{V_{\text{Man},t}}
{V_{\text{Man},t} + V_{\text{Semi},t} + V_{\text{Raw},t}} \times 100
$$
 
The share rises either when the numerator grows or when
$V_{\text{Raw},t}$ falls, and shares alone cannot tell the two apart.
The absolute values settle it.
 
```{r}
yearly = aggregate(value ~ cat + year, data = monthly, FUN = sum)
yearly[yearly$year %in% c(2015, 2025), ]
```
 
 The absolute figures are decisive, and they are more favourable to the
diversification story than the shares alone suggest. Manufactured exports
rose from **124.6** to **340.1** billion riyals — a factor of **2.7**, or
**10.6%** a year compounded. Crucially, raw material exports did **not**
collapse: they grew from **501.8** to **637.7** billion, up **27%** over
the decade. Semi-manufactured exports grew **40%**.

All three categories expanded. The manufactured share therefore rose not
because oil exports shrank, but because manufacturing grew roughly four
times faster than raw materials — **10.6%** a year against **2.4%**.
This is the strongest form the result could take: a compositional shift
driven entirely by differential growth, with no contraction anywhere in
the denominator.

## 4.3 Testing the trend
 
A rising line in a chart is not evidence on its own. We fit a linear
trend and test whether its slope differs from zero:
 
$$
s_t = \beta_0 + \beta_1 t + \varepsilon_t,
\qquad
H_0: \beta_1 = 0
\quad \text{vs} \quad
H_1: \beta_1 \neq 0
$$
 
where $\beta_1$ is the average change in the share per month, in
percentage points.
 
```{r}
trend_model = lm(share ~ t, data = man)
summary(trend_model)
```
 
The ordinary least squares standard errors above are not trustworthy on
their own. In a time series the residuals are **autocorrelated**: a month
that sits above the trend line is usually followed by another month above
it. OLS treats the observations as independent, so it understates the
standard error and overstates significance. The Newey-West estimator
corrects for this, leaving the slope estimate unchanged.
 
```{r}
coeftest(trend_model, vcov = NeweyWest(trend_model))
```
The slope survives the correction. At **0.079** percentage points a month
— **0.95** points a year — the trend remains significant at the 1% level
($p = 0.0034$) even with the standard error widened for autocorrelation.
The linear model explains **55%** of the variation in the manufactured
share ($R^2 = 0.554$), which is high for a monthly series of this length
and confirms that the movement in the chart is a trend rather than drift
around a fixed level.
 
```{r}
# Slope in percentage points per year
round(coef(trend_model)["t"] * 12, 3)
```
 
# 5. Structure of the Series
 
```{r}
man_ts = ts(man$share, start = c(2015, 1), frequency = 12)
 
length(man_ts)      # 137
start(man_ts)       # 2015/1
end(man_ts)         # 2026/5
frequency(man_ts)   # 12
```
 
## 5.1 STL Decomposition
 
STL splits the series into three additive components:
 
$$
y_t = T_t + S_t + R_t
$$
 
the trend-cycle, the seasonal component, and the remainder.
 
```{r, fig.width=8, fig.height=5}
fit = stl(man_ts, s.window = "periodic")
 
autoplot(fit) +
  labs(title = "STL Decomposition of the Manufactured Share")
```
 
## 5.2 Strength of Each Component
 
A common mistake here is to divide each component's variance by the
variance of the series and read the results as percentages. The
components are correlated, so those figures do not add to 100 and cannot
be interpreted as shares. The standard measure compares each component
against the remainder instead, giving a value between 0 and 1:
 
$$
F_T = \max\!\left(0,\, 1 - \frac{\mathrm{Var}(R_t)}{\mathrm{Var}(T_t + R_t)}\right),
\qquad
F_S = \max\!\left(0,\, 1 - \frac{\mathrm{Var}(R_t)}{\mathrm{Var}(S_t + R_t)}\right)
$$
 
A value near 1 means the component dominates the noise; a value near 0
means it is indistinguishable from it.
 
```{r}
comp = fit$time.series
R    = comp[, "remainder"]
round(c(trend    = max(0, 1 - var(R) / var(comp[, "trend"]    + R)),
        seasonal = max(0, 1 - var(R) / var(comp[, "seasonal"] + R))), 3)
 
 
round(c(seasonal  = diff(range(comp[, "seasonal"])),
        remainder = diff(range(R)),
        data      = diff(range(man_ts))), 2)
```
The decomposition is lopsided. The trend strength is **0.883**, close to
the upper bound of 1: almost everything in this series is trend. The
seasonal strength is **0.082**, near zero, so the seasonal component is
barely distinguishable from noise.

The ranges confirm it. The seasonal component spans only **1.57**
percentage points across the whole period, against **9.26** for the
remainder and **20.20** for the series itself. Seasonality accounts for
less than a tenth of the movement the data actually shows — a finding
that shapes the modelling choices in Section 6.



## 5.3 Seasonality Check
 
```{r}
a = acf(man_ts, lag.max = 24, main = "ACF - Seasonal Check")
setNames(round(as.numeric(a$acf)[12:14], 3), paste("lag", 11:13))
round(2 / sqrt(length(man_ts)), 3)
```

There is no annual seasonality. A seasonal series would show an isolated
peak at lag 12, standing clear of its neighbours. Instead the three
values decline smoothly — **0.453** at lag 11, **0.431** at lag 12,
**0.413** at lag 13 — with no peak at all.

All three exceed the significance bound of **0.171**, but that reflects
the trend rather than a seasonal cycle: a trending series is correlated
with itself at every lag. The gradual decay is the signature of a trend,
not of a repeating annual pattern, and it agrees with the seasonal
strength of 0.082 found above. 
 
 
## 5.4 Shock Analysis
 
```{r}
idx = order(abs(R), decreasing = TRUE)[1:6]
 
data.frame(date  = format(man$date[idx], "%Y-%m"),
           shock = round(R[idx], 2))
```
 
```{r}
monthly[monthly$date %in% man$date[idx],
        c("date", "cat", "value", "share")]
```


Two of these months make the same point from opposite directions, and
together they justify the whole approach of this report.

**July 2025** is the largest positive shock at **+5.22** points. Here the
share of 33.6% reflects genuine expansion: manufactured exports reached
**34,256** million riyals, their highest value in the entire series,
while raw materials at 51,432 million sat near their long-run average.
The numerator moved.

**May 2020** is the fourth largest at **+3.45** points — and it means the
opposite. Manufactured exports that month were only **8,182** million,
close to their minimum for the whole period. The share rose because raw
materials collapsed to **19,859** million, their lowest value in eleven
years, as oil markets seized during the first months of the pandemic. The
denominator moved.

Read from the shares alone, both months look like progress. Read against
the absolute values, one is growth and the other is contraction elsewhere.
This is precisely the ambiguity set out in Section 4.2, appearing here in
two specific months.

The negative shocks share a pattern: February 2020, October and November
2023, and January 2017 are all months when raw materials stood well above
their average — between 66.9% and 69.6% of exports — compressing the
manufactured share arithmetically. 
 
 
 
# 6. Modelling
 
## 6.1 Stationarity Test
 
ARIMA requires a stationary series. The Augmented Dickey-Fuller test
takes
 
$$
H_0: \text{the series has a unit root (non-stationary)}
$$
 
Note that the null here is the *undesirable* case, the reverse of the
usual convention: a **small** $p$-value is the good outcome.
 
```{r, warning = FALSE}
adf.test(man_ts)
adf.test(diff(man_ts))
```
The result is unambiguous. On the level the test statistic is **-1.08**
with $p = 0.92$: nowhere near rejection, so we cannot rule out a unit
root. After first differencing the statistic falls to **-6.23** with
$p < 0.01$, comfortably rejecting non-stationarity.

The series is therefore integrated of order one, $d = 1$, and the
differenced series is what the model should be built on. 
 
 
 
## 6.2 ACF and PACF After Differencing
 
```{r, fig.width=8, fig.height=3.2}
p1 = ggAcf(diff(man_ts),  lag.max = 24) + labs(title = "ACF")
p2 = ggPacf(diff(man_ts), lag.max = 24) + labs(title = "PACF")
 
p1 | p2
```

The two functions point the same way. The ACF of the differenced series
shows significant **negative** autocorrelation at lags 1 and 2, both
beyond the confidence bounds, and then falls inside them and stays there.
The PACF, by contrast, does not cut off: it starts negative and decays
gradually across the early lags.

That combination is the signature of a moving-average process of order
two. An $\mathrm{AR}(p)$ process would show the reverse — a PACF cutting
off at lag $p$ with a slowly decaying ACF. The reading therefore suggests
$\mathrm{ARIMA}(0,1,2)$, and Section 6.3 can be checked against it rather
than accepted on trust.

Neither plot shows anything notable at lag 12, which agrees with the
seasonality check in Section 5.3. A small number of bars near lags 20 to
23 touch the bounds, but with 24 lags examined at the 5% level one or two
such crossings are expected by chance and no isolated lag carries a
structural interpretation.


 
## 6.3 Model Selection
 
The general model is
 
$$
\phi(B)(1 - B)^d y_t = c + \theta(B)\,\varepsilon_t
$$
 
where $B$ is the backshift operator and $\phi$, $\theta$ are the
autoregressive and moving-average polynomials of orders $p$ and $q$. We
hold out the last twelve months so the model can later be judged on data
it has never seen.
 
```{r}
train = window(man_ts, end = c(2025, 5))
 
fit1 = auto.arima(train)
summary(fit1)
```
 
## 6.4 Comparing Alternatives
 
AIC is only comparable across models fitted to the **same** differenced
series. If `auto.arima` selected a different order of differencing, or
added a seasonal part, the comparison below is meaningless. So we check
the order first.
 
```{r}
setNames(fit1$arma, c("p", "q", "P", "Q", "period", "d", "D"))
```
 
```{r}
stopifnot(fit1$arma[6] == 1, fit1$arma[3] == 0, fit1$arma[4] == 0)
 
fit2 = Arima(train, order = c(2, 1, 0))
fit3 = Arima(train, order = c(2, 1, 2))
 
round(c(auto = fit1$aic, ar2 = fit2$aic, both = fit3$aic), 2)
```


 The automatic selection wins clearly. ARIMA(0,1,2) with drift scores
**480.08**, against **484.99** for ARIMA(2,1,2) and **486.24** for
ARIMA(2,1,0). A gap of five to six AIC points is substantial. The
agreement matters as much as the ranking: the specification chosen by
AICc is the same one the correlation plots pointed to, so two independent
routes — visual identification and automatic search — converge on the
same model.



# 7. Validation
 
## 7.1 Out-of-Sample Accuracy
 
With $e_t$ the forecast errors over the $h = 12$ held-out months,
 
$$
\mathrm{MASE} =
\frac{\frac{1}{h}\sum |e_t|}
     {\frac{1}{n-1}\sum_{t=2}^{n} |y_t - y_{t-1}|}
$$
 
The denominator is the error of a naive forecast $\hat{y}_t = y_{t-1}$,
so $\mathrm{MASE} < 1$ means the model beats carrying the last value
forward. The **Test set** row is the one that matters.
 
```{r}
test = window(man_ts, start = c(2025, 6))   # 12 months the model never saw
 
fc = forecast(fit1, h = 12)
 
round(accuracy(fc, test), 3)
```

The model forecasts better than the benchmark. On the twelve held-out
months the MASE is **0.723**, meaning the average error is about 72% of
what a naive "carry the last value forward" forecast would produce.
Theil's U of **0.916** points the same way. The mean absolute percentage
error is **5.8%**.

One feature deserves attention. The mean error on the test set is
**+1.55**, and a positive mean error means the model **under-forecast**
the manufactured share in those months: the rise through late 2025 and
early 2026 was faster than the fitted drift anticipated. The gap between
the training MASE of 0.523 and the test MASE of 0.723 is the expected
cost of forecasting genuinely unseen data, and the size of that gap is
modest.

 
## 7.2 Residual Diagnostics
 
If the model has captured the structure of the series, nothing
predictable should remain in the residuals. The Ljung-Box test takes
independence as its null, so here a **large** $p$-value is the good
result.
 
```{r}
checkresiduals(fit1)
```
 The residuals pass. The Ljung-Box statistic over 24 lags is **19.09** on
22 degrees of freedom, giving $p = 0.64$ — far from rejection. There is
no evidence of remaining autocorrelation, the residual plot shows no
systematic pattern, and the histogram is approximately normal.

This does not prove the model is correct; a test that fails to reject is
not a test that confirms. What it does establish is that no obvious
structure has been left unmodelled.

# 8. Forecast
 
## 8.1 Final Model
 
```{r}
fit_final = auto.arima(man_ts)
summary(fit_final)
```
Refitting on all 137 observations selects **the same specification** as
the model validated in Section 7 — ARIMA(0,1,2) with drift — so the
out-of-sample performance reported there applies directly to the model
used for forecasting. The coefficients shift slightly with the extra
year of data, and the drift term strengthens from 0.100 to **0.1125**
percentage points a month, now significant at the 5% level. The recent
months have reinforced the upward trend rather than tempering it.

## 8.2 Forecast to 2030
 
The horizon runs from June 2026 to December 2030, $h = 55$ months.
Prediction intervals take the form
 
$$
\hat{y}_{T+h} \pm z_{1-\alpha/2}\,\hat{\sigma}_h,
\qquad \hat{\sigma}_h \propto \sqrt{h}
$$
 
which is why the band widens as it moves right rather than staying
parallel.
 
```{r, fig.width=8, fig.height=4}
fc_2030 = forecast(fit_final, h = 55)
 
autoplot(fc_2030) +
  geom_hline(yintercept = tail(fc_2030$mean, 1),
             linetype = "dashed", color = "grey50", linewidth = 0.4) +
  labs(title = "Forecast of the Manufactured Share to 2030",
       subtitle = paste0(as.character(fit_final),
                         ", 80% and 95% prediction intervals"),
       x = NULL, y = "Share of total exports (%)") +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(color = "grey40", size = 9.5))
```
 
```{r}
round(c(forecast = tail(fc_2030$mean, 1),
        lower95  = tail(fc_2030$lower[, 2], 1),
        upper95  = tail(fc_2030$upper[, 2], 1)), 2)
```
 
By December 2030 the model places the manufactured share at
**`r sprintf("%.1f%%", tail(fc_2030$mean, 1))`**, with a 95% prediction
interval running from **`r sprintf("%.1f%%", tail(fc_2030$lower[, 2], 1))`**
to **`r sprintf("%.1f%%", tail(fc_2030$upper[, 2], 1))`**. The width of
that interval is itself part of the result, and Section 10 explains why
it should not be read as a precise target.
 
# 9. Summary
 
> **The composition has shifted.** Manufactured goods made up
> **`r sprintf("%.0f%%", man$share[1])`** of Saudi exports at the start of
> 2015 and **`r sprintf("%.0f%%", man$share[nrow(man)])`** by mid-2026.
> Every category grew in absolute terms over the decade — oil exports
> included — but manufactured exports grew more than four times faster
> than raw materials, and that difference is what moved the structure.

## 9.1 The three categories
 
```{r, echo=FALSE}
yr = function(v, y, f) tapply(v[monthly$year == y], monthly$cat[monthly$year == y], f)
 
s15 = yr(monthly$share, 2015, mean);  s25 = yr(monthly$share, 2025, mean)
a15 = yr(monthly$value, 2015, sum);   a25 = yr(monthly$value, 2025, sum)
 
cat_tbl = data.frame(
  Category                  = names(s15),
  `Share in 2015`           = sprintf("%.1f%%", s15),
  `Share in 2025`           = sprintf("%.1f%%", s25),
  `Change`                  = sprintf("%+.1f points", s25 - s15),
  `Growth in value, a year` = sprintf("%+.1f%%", 100 * ((a25 / a15)^(1/10) - 1)),
  check.names               = FALSE)
 
kbl(cat_tbl, align = c("l", "r", "r", "r", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
 
The last column is the one that settles the question. A share can rise
simply because something else fell, but this column measures each
category against its own past, in riyals. Manufactured exports grew on
their own terms.
 
## 9.2 The figures at a glance
 
```{r, echo=FALSE}
plain = data.frame(
  Quantity = c(
    "Period covered",
    "Total exports over the period",
    "Average exports a year",
    "Manufactured share, start",
    "Manufactured share, latest",
    "Change over the period",
    "Average change a year",
    "Projected share by December 2030",
    "Range the projection could fall in"),
  Value = c(
    paste(format(min(monthly$date), "%B %Y"), "to",
          format(max(monthly$date), "%B %Y")),
    sprintf("%.1f trillion riyals", sum(monthly$value) / 1e6),
    sprintf("%.2f trillion riyals", 12 * mean(tot) / 1e6),
    sprintf("%.1f%%", man$share[1]),
    sprintf("%.1f%%", man$share[nrow(man)]),
    sprintf("%+.1f points", man$share[nrow(man)] - man$share[1]),
    sprintf("%+.2f points", coef(trend_model)["t"] * 12),
    sprintf("%.1f%%", tail(fc_2030$mean, 1)),
    sprintf("%.0f%% to %.0f%%", tail(fc_2030$lower[, 2], 1),
                                tail(fc_2030$upper[, 2], 1))))
 
kbl(plain, col.names = c("", "Value"), align = c("l", "r")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
 
The last row deserves emphasis. The projection is a direction, not a
target: the range around it is wide because four and a half years is a
long way to look ahead from eleven years of data.
 
# 10. Limitations
 
Four constraints affect how the results above should be read.
 
1. **Shares are not independent.** The adding-up constraint implies
 
   $$
   \sum_i \Delta s_{i,t} = 0
   $$
 
   so the three series cannot move freely. A rise in the manufactured
   share is arithmetically identical to a fall in the sum of the other
   two. This is why Section 4.2 checks the absolute values: it is the
   only part of the analysis that can distinguish real growth in
   manufacturing from a mechanical effect of falling oil exports.
 
2. **The forecast horizon is long relative to the data.** We project 55
   months forward from 137 observations. Prediction intervals widen with
   the square root of the horizon, which is why the 2030 interval is so
   much wider than the 2027 one. The central forecast is a continuation
   of the estimated drift, not a prediction of policy.
 
3. **The model can leave the admissible range.** A share is bounded
   between 0 and 100, but an ARIMA model with drift extrapolates linearly
   and knows nothing about that bound. In this case the upper 95% limit
   reaches 46.3% and the problem does not bind, but that is a property of
   these particular numbers rather than of the method. Modelling

   $$
   z_t = \log\!\left(\frac{s_t}{100 - s_t}\right)
   $$

   and inverting afterwards would respect the bound by construction, at
   any horizon. We did not apply it here because the raw limits remained
   plausible.
 
4. **Structural breaks are treated as noise.** The shocks identified in
   Section 5.4 — the 2020 collapse in particular — are absorbed into the
   remainder rather than modelled. A model that allowed the trend to
   change slope after 2020 might fit the recent years better, at the cost
   of relying on a break date chosen after seeing the data.
 
# 11. Conclusion
 
The question was whether the composition of Saudi exports has shifted
from raw materials toward manufactured goods between 2015 and 2026. The
evidence points in one direction, with qualifications.
 
The manufactured share rose from
**`r sprintf("%.1f%%", man$share[1])`** to
**`r sprintf("%.1f%%", man$share[nrow(man)])`** over the period, an
average of **`r sprintf("%.2f", coef(trend_model)["t"] * 12)`** percentage
points a year. The trend remains statistically significant after
correcting the standard errors for autocorrelation, so it is not a
by-product of month-to-month noise.
 
Section 4.2 shows there was no denominator effect at all. Raw material
exports grew 27% over the decade and semi-manufactured exports 40%, while
manufactured exports grew 173% — a compound rate of 10.6% a year against
2.4% for raw materials. The shift in composition came entirely from
differential growth, not from contraction anywhere.
 
Extending the fitted model to 2030 gives a central estimate of
**`r sprintf("%.1f%%", tail(fc_2030$mean, 1))`**. The prediction interval
is wide, and the limitations in Section 10 explain why: the horizon is
long, the model is unaware that a share is bounded, and the 2020 break is
absorbed rather than modelled. The direction of the result is firmer than
its magnitude.
 
# 12. Technical Appendix
 
The figures below are for readers who want to check the statistical basis
of the results. Each is explained in the section it came from.
 
```{r, echo=FALSE}
Ft = max(0, 1 - var(R) / var(comp[, "trend"]    + R))
Fs = max(0, 1 - var(R) / var(comp[, "seasonal"] + R))
nw = coeftest(trend_model, vcov = NeweyWest(trend_model))
lb = Box.test(residuals(fit1), lag = 24,
              fitdf = sum(fit1$arma[1:2]), type = "Ljung-Box")
 
tech = data.frame(
  Quantity = c(
    "Observations",
    "Trend slope (percentage points per year)",
    "Trend p-value, Newey-West corrected",
    "Trend strength, F_T",
    "Seasonal strength, F_S",
    "Selected model",
    "Out-of-sample MASE (12 months held out)",
    "Ljung-Box p-value, 24 lags",
    "Forecast horizon"),
  Value = c(
    sprintf("%d months x 3 categories", nrow(man)),
    sprintf("%+.3f", coef(trend_model)["t"] * 12),
    format.pval(nw["t", 4], digits = 3, eps = 1e-4),
    sprintf("%.3f", Ft),
    sprintf("%.3f", Fs),
    as.character(fit_final),
    sprintf("%.3f", accuracy(fc, test)["Test set", "MASE"]),
    sprintf("%.3f", lb$p.value),
    "55 months"),
  Section = c("2", "4.3", "4.3", "5.2", "5.2", "8.1", "7.1", "7.2", "8.2"))
 
kbl(tech, align = c("l", "r", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
 
# 13. References
 
General Authority for Statistics (GASTAT). *International Trade
Statistics — Exports by Nature of Materials*, monthly series,
January 2015 – May 2026. Statistical Database.
**Accessed 17 September 2026.**